\chapter{Introduction}
\label{chapter:introduction}

%% Target length: ~4 pages
%%
%% 1.1  Motivation
%%      - Live-cell imaging: spatial resolution vs. temporal resolution vs.
%%        sample health trade-off (cite Pylvänäinen 2023).
%%      - Diffraction barrier ~200-300 nm.
%%      - SOFI: statistical SR from blinking fluorophores; needs hundreds of
%%        frames → limits temporal resolution.
%%      - Deep learning potential to reduce frame count (bridge to DL MISR).
%%
%% 1.2  Prior Work in the Group
%%      - Jelle Komen (2024): SOFI-MISRGRU — ConvGRU fusion achieved
%%        real-time SR at up to 4.85 fps using 20 frames. Establishes the
%%        baseline this thesis builds upon.
%%      - Gap: recurrent processing is sequential; Transformers with
%%        self-attention process all frames in parallel and may capture
%%        richer long-range correlations across the fluorophore blinking
%%        sequence.
%%
%% 1.3  Research Questions
%%      Q1: Can a Transformer-based MISR architecture (TR-MISR) replace the
%%          ConvGRU fusion stage and maintain or improve SR quality on SOFI?
%%      Q2: Which decoder variant (PixelShuffle vs. ConvTranspose2d) better
%%          matches the 2n−7 SOFI output size requirement?
%%      Q3: What is the effect of different Fourier-domain loss functions on
%%          high-frequency reconstruction?
%%      Q4: Does the Transformer's parallel processing enable lower inference
%%          latency than the recurrent baseline after TensorRT optimisation?
%%
%% 1.4  Contributions
%%      1. Adaptation of TR-MISR to single-channel fluorescence microscopy
%%         with SOFI-specific pre-processing and patch sizes.
%%      2. Systematic comparison of TRNet, TRNet_deconv, and TRNet_improved
%%         (FlashAttention + CrossAttentionAggregator) against SOFI-MISRGRU.
%%      3. Ablation study: loss functions (L1, Fourier, L1_Fourier, Composite),
%%         LR scheduler strategies, training set size effects.
%%      4. TensorRT inference benchmarking and speed-accuracy trade-off analysis.
%%      5. (If applicable) Streaming MISRGRU: N-input → N-output hidden-state
%%         architecture for rolling-window live-cell imaging.
%%
%% 1.5  Thesis Outline
%%      One sentence per chapter.


Capturing dynamic cellular interactions is a fundamentally difficult problem due to maximizing spatial and temporal resolution while minimizing sample damage. Electron microscopes can achieve sub nanometer resolution, however, that requires to first kill the sample, thus no interactions and movement cannot be observed. The optical microscope is limited by the diffraction limit. To improve upon spatial resolution and see past the diffraction limit several techniques have been developed. They all rely on fluorescent molecules, called fluorophores, which are first used to label the specime. They bind to target proteins or structures and emit light upon excitation. A sequence of frames is then captured, with molecules randomly blinking across the frames. Different fluorophores can blink either sparsely for precise localization or densely. 


Super-resolution (SR) techniques can overcome this barrier. Well-known methods include Stimulated Emission Depletion (STED) microscopy \ref{}, Structured Illumination Microscopy (SIM) \ref{}, and Single Molecule Localization Microscopy (SMLM) \ref{}, which can achieve spatial resolution down to 10 nm. However, these methods either require complex microscope setups, high signal-to-noise ratio (SNR) conditions, or millions of frames to reconstruct a single SR image, all of which limit temporal resolution and risk sample damage \ref{}. Other techniques, such as Super-resolution Optical Fluctuation Imaging (SOFI) \ref{} and Super-Resolution Radial Fluctuations (SRRF) microscopy \ref{}, do not require specialized hardware and can typically operate with only hundreds of frames, accommodating dense fluorophore blinking.

Among these, SOFI is uniquely grounded in statistical theory: it reconstructs SR images by exploiting the temporal and spatial correlations of fluorophore blinking through cross-cumulant analysis. Resolution improvement arises from the statistical independence of different fluorophores \ref{}. Because noise is temporally uncorrelated, SOFI can operate under low SNR conditions, making it particularly well-suited for live-cell imaging. For n-th order SOFI, an n-fold improvement in spatial resolution is achieved. Despite these advantages, SOFI still requires hundreds of frames per reconstruction, which limits its ability to capture fast cellular dynamics and renders real-time imaging impractical. Without real-time capability, researchers cannot make immediate decisions during live experiments, reducing the practical utility of the technique.                                
Deep learning offers a promising path to reduce the number of frames required for SR reconstruction while preserving resolution gains. Recent work by TODO JELLE\ref{} introduced a deep learning model that accelerates second-order SOFI by reconstructing a 2× SR image from as few as 20 frames. Their model encodes input frames into feature maps, fuses them with a recurrent Gated Recurrent Unit (GRU) structure to capture blinking correlations across time, and upsamples the result into a second-order SR image. Trained on synthetic data the approach demonstrated that spatiotemporal deep learning models can approximate SOFI-quality reconstructions at a fraction of the acquisition cost.
        
\todo[inline]{mention streaming architecture}
In this thesis, we investigate whether the recurrent fusion module can be replaced by a more expressive attention-based mechanism. Specifically, we adapt TR-MISR \ref{} a Transformer-based Multi-Image Super-Resolution framework originally developed for satellite imagery and ranked first on the PROBA-V benchmark, to the SOFI acceleration task. Rather than fusing frames sequentially through a GRU, TR-MISR employs a multi-head self-attention. Transformer with learnable class tokens to aggregate spatial features across all input frames simultaneously.                                                                                                                    
\todo[inline]{references}
\todo[inline]{mention streaming architecture}
\todo[inline]{results}


\section{Outline}
In the next chapters, the concepts such as fluorescent microscopy and deep learning are presented in chapter \ref{ch:background}. Chapter \ref{ch:method} presents our proposed architecture and how we augmented the datasets for moving structures followed by chapter \ref{ch:results}. Lastly, the main outcomes are provided in chapter \ref{ch:results}.
\emph{Introduction.. \cite{example-article}}