\chapter{TRNet Architecture and Training Ablations}
\label{app:trnet_modifications}

This appendix reports the ablation experiments that informed the choice of decoder and
loss function used in all main-chapter TR-MISR experiments
(Section~\ref{ssec:experimental_setup}).
Two design choices were evaluated independently against the chosen baseline:
First, replacing the PixelShuffle decoder with a transposed-convolution (deconvolution) decoder,
and secondly, replacing the combined L1+Fourier loss with the Fourier-only loss.
All other hyperparameters are held fixed at the values described in
Section~\ref{ssec:experimental_setup}.

%% -------------------------------------------------------
\section{Decoder: PixelShuffle vs.\ Transposed Convolution}
\label{app:decoder_ablation}

The SOFI super-resolution task requires upscaling LR from $K \times K$ to HR result of size 
$2*K-7 \times 2*K-7$.
Two standard decoder approaches were compared:

\begin{itemize}
  \item \textbf{PixelShuffle} (sub-pixel convolution~\cite{shi2016pixelshuffle}):
   Originaly used in the TR-MISR paper~cite\cite{TR-MISR-original}. It learns
    to rearrange channel dimensions into spatial resolution. The non-integer scale is handled
    by upsampling to an intermediate size followed by a centre crop to match the $2K-7$.
  \item \textbf{Transposed convolution} (\texttt{ConvTranspose2d}): Is used in MISRGRU models from 
  RESURF~\cite{resurf2024} learned upsampling via
    strided deconvolution. Known to produce checkerboard artefacts when the kernel size is
    not divisible by the stride~\cite{odena2016deconvolution}.
\end{itemize}

%% -------------------------------------------------------
\section{Loss Function: L1+Fourier vs.\ Fourier Only}
\label{app:loss_ablation}

The Fourier loss~\cite{resurf2024} penalises errors in the amplitude and phase spectra of
the prediction, encouraging recovery of high-frequency detail that an L1 loss would
suppress via mean regression. A combined L1+Fourier loss retains the spatial stability of
L1 while adding spectral supervision. The ablation tests whether the L1 term is redundant
or provides complementary signal.

%% -------------------------------------------------------
\section{Results}
\label{app:ablation_results}

Table~\ref{tab:ablation_aggregate} reports aggregate metrics over all 480 test samples
(12 SNR conditions $\times$ 40 samples each). Table~\ref{tab:ablation_per_snr} gives
the RSP breakdown per SNR condition.

\begin{table}[h]
  \centering
  \caption{Aggregate metrics for the three configurations over 480 test samples.
           Bold marks the best value per column.}
  \label{tab:ablation_aggregate}
  \begin{tabular}{lccccc}
    \toprule
    \textbf{Model} & \textbf{RSP} & \textbf{PSNR (dB)} \\
    \midrule
    L1+Fourier, PixelShuffle (baseline) & \textbf{0.8789} & \textbf{23.54} \\
    Fourier only, PixelShuffle          & 0.8757          & 23.08          \\
    L1+Fourier, Deconv                  & 0.8770          & 23.43          \\
    \bottomrule
  \end{tabular}
\end{table}

\begin{table}[h]
  \centering
  \caption{RSP per SNR condition. Conditions are ordered by blinking speed
           (fast/medium/slow) and photon level (100\%--40\%). Bold marks the best
           value per row.}
  \label{tab:ablation_per_snr}
  \begin{tabular}{lccc}
    \toprule
    \textbf{Condition} & \textbf{L1+F, PS (baseline)} & \textbf{F only, PS} & \textbf{L1+F, Deconv} \\
    \midrule
    01 fast p100  & \textbf{0.9013} & 0.8987 & 0.8995 \\
    02 fast p80   & \textbf{0.8996} & 0.8984 & 0.8988 \\
    03 fast p60   & 0.9022          & \textbf{0.9024} & 0.9021 \\
    04 fast p40   & 0.8950          & \textbf{0.8998} & 0.8964 \\
    \midrule
    05 medium p100 & \textbf{0.8885} & 0.8881 & 0.8872 \\
    06 medium p80  & 0.8900          & \textbf{0.8915} & 0.8892 \\
    07 medium p60  & 0.8845          & \textbf{0.8872} & 0.8839 \\
    08 medium p40  & 0.8843          & \textbf{0.8861} & 0.8842 \\
    \midrule
    09 slow p100   & \textbf{0.8579} & 0.8548 & 0.8545 \\
    10 slow p80    & \textbf{0.8536} & 0.8483 & 0.8498 \\
    11 slow p60    & \textbf{0.8556} & 0.8453 & 0.8510 \\
    12 slow p40    & \textbf{0.8345} & 0.8073 & 0.8269 \\
    \midrule
    \textbf{Aggregate} & \textbf{0.8789} & 0.8757 & 0.8770 \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Decoder comparison}

The Deconv decoder performs comparably to PixelShuffle at high-SNR conditions (fast
blinking, high photon counts) but falls behind consistently at medium and slow blinking.
At the most challenging condition (12\_slow\_p40), the PixelShuffle baseline achieves
RSP~$=0.835$ versus $0.827$ for Deconv, a gap of $0.008$ that represents a meaningful
difference in structural fidelity under low-SNR conditions. The aggregate PSNR difference
is smaller (23.54 vs.\ 23.43\,dB), suggesting the gap is primarily in structural quality
rather than pixel-level intensity accuracy.

The PixelShuffle advantage likely arises from the non-integer upscaling factor
($64 \to 121$). Transposed convolution with non-divisible stride/kernel produces periodic
checkerboard artefacts~\cite{odena2016deconvolution}, which manifest as spurious
high-frequency patterns in the SR output. These artefacts are not strongly penalised by
the Fourier amplitude loss (which averages over all frequencies) but do degrade RSP
because they introduce spatial inaccuracies that do not match the ground-truth SOFI.
PixelShuffle avoids this by working entirely in the channel domain and cropping to the
target size, producing smooth outputs without periodic artefacts.

\subsection{Loss function comparison}

The Fourier-only loss matches or slightly exceeds the baseline at fast-blinking, high-SNR
conditions (conditions 03--04 and 06--08) but degrades substantially at slow-blinking,
low-photon conditions. At condition 12\_slow\_p40, the Fourier-only model achieves
RSP~$=0.807$ versus $0.835$ for the baseline, a gap of $0.027$.
 The gap between baseline and Fourier-only grows with decreasing SNR.

The L1 term therefore provides complementary supervision: it penalises spatial
misregistration directly, stabilising the model under challenging imaging conditions
where Fourier-based supervision alone is insufficient.

%% -------------------------------------------------------

\subsection{Visual inspection}
\todo{compare TRMISR contrast next to lines artefact with Fourier one which doenst have this}
\todo{ask miyase how to get tif files into thesis as png quality is shitty}

\section{Summary}

\begin{itemize}
  \item \textbf{PixelShuffle decoder} for the TRNet is preferred over transposed convolution  
  as it produces consistently higher RSP, especially at low SNR.
  \item \textbf{L1+Fourier loss} is preferred over Fourier-only for TRNet models. 
  The L1 term provides 
  spatial regularisation that prevents the model from placing structures at incorrect
    positions, with the benefit growing under challenging (slow-blinking, low-photon)
    conditions.
\end{itemize}

These findings motivated the use of the PixelShuffle decoder and L1+Fourier loss in all
TR-MISR experiments reported in Chapter~\ref{ch:results}.
